---
title: "Awesome 3D Interactive Globe Tutorial"
author: "Caio Lente"
date: "2017-02-08"
output: 
  html_document: 
    fig_height: 7
    fig_width: 10
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I'm addicted to spatial visualizations; I think they really put datasets into perspective, 
as if each data point has its place. In this tutorial I'm going to teach you how to create a 
very cool 3D interactive globe with `threejs`.

Our goal will be to visualize Japan's commerce with arcs on a 3D earth. The source dataset 
we'll be using ([Japan Trade Statistics](https://www.kaggle.com/zanjibar/japan-trade-statistics/kernels) by 
[Tadashi Nagao](https://www.kaggle.com/zanjibar)) is a really good example of what you'll find in the wild: 
the data only talks about countries (with their names recorded in a non-standard way) and has absolutely no 
information about their coordinates.

To create the visualization there are 3 steps: geting the coordinates of the capital of each country, 
converting the country names to a more standard form and creating the globe itself.

## Getting the coordinates

```{r include=FALSE, message=FALSE}
# Load packages
library(tidyverse)
library(stringr)
library(threejs)
library(maps)

# Load data
countries <- read_csv("Data/country_eng.csv")
trades <- read_csv("Data/ym_latest.csv")

# Make names of columns consistent
names(countries) <- str_to_lower(names(countries))
names(trades) <- str_to_lower(names(trades))

# Join tables and keep only relevant columns
trades <- trades %>%
  left_join(countries) %>%
  select(country_name, country, value) %>%
  group_by(country_name, country) %>%
  summarise(value = sum(as.numeric(value))) %>%
  ungroup()
```

Our first task is importing the data.

I loaded the table with Japan's trade information, loaded the table with the country/country-code 
relationship, and then joined the two into a `trades` data frame. I also converted the column names to 
lower case, selected only the relevant variables, and grouped the table by country (adding up the 
values of all trades with each country). None of this is shown here because it is too dataset-specific, 
but this was the result:

```{r, echo=FALSE}
knitr::kable(head(trades))
```

After this initial setup, we can move on to getting the actual coordinates of the capital of each country 
with the `maps` package (we're going to use the coordinates of the capitals as the landing points for 
the arcs on the 3D globe). Here I'll load the `world.cities` dataset and select the names of the countries, 
their capitals, and their coordinates.

```{r}
# Get coordinates of every capital
capitals <- world.cities %>%
  filter(capital == 1) %>%
  transmute(
    country_name = country.etc,
    lat = lat,
    long = long
  ) %>%
  rbind(c("Hong Kong", 22.39, 114.1))
```

**Note**: Let's not get caught up in geopolitics, but I added Hong Kong manually because it apparently 
is/isn't part of China, so it isn't in the `world.cities` dataset...

This is what `capitals` looks like:

```{r, echo=FALSE}
knitr::kable(head(capitals))
```

## Converting country names

Take a look at how "American Samoa" and "United Arab Emirates" are spelled in the two tables above 
(that's the main reason why I included them). They are not formated in the same way.

I'll spare you the details, but this is the most difficult taks you'll face when dealing with spatial 
data: names are a pain. The two tables we have to work with encode the country names in very different 
ways, so we need to standarize those.

We'll do this with the `countrycodes` package, more specifically the `countrycode_data` dataset, which 
contains a column of regex statements to match almost any way you could write the name of each country. 
First I clean just a little bit the names in `trades$country_name`, but then it's as straight forward as this:

```{r, include=FALSE}
# Remove underscores, remove country abreviations in parentheses, and
# change South Korea's name so that it matches
clean_names <- function(names) {
  names <- str_replace_all(names, "_", " ")
  names <- str_replace_all(names, "\\([a-zA-Z]+\\)", "")
  names[str_detect(names, "Korea")] <- "South Korea"
  
  return(names)
}

# Find the entry corresponding to 'regex[i]' in 'names' and return a
# vector with the matches
match_country <- function(regex, names) {
  matches <- c()
  for (i in 1:length(regex)) {
    match <- grep(regex[i], names, ignore.case = TRUE, perl = TRUE)
    matches[i] <- ifelse(length(match) == 0, NA, names[match])
  }
  
  return(matches)
}
```

```{r, include=FALSE}
# Clean the country names in 'trades'
trades$country_name <- clean_names(trades$country_name)
```

```{r}
# Get the regex statement for each country
regex <- countrycode::countrycode_data$country.name.en.regex

# Get a correspondence between country names in 'totals' and 'capitals'
trades_capitals <- as_tibble(cbind(
  regex,
  trades_countries = match_country(regex, trades$country_name),
  capitals_countries = match_country(regex, capitals$country_name)
))
```

On the snippet above I'm omitting the `match_country` function, which simply matches each entry in 
`regex` with an entry of the second argument, but the overall code is pretty simple! And now we have 
a table that connects the names in `trades` with the names in `capitals`, which we can use to join 
the two other tables into a new one called `geo_trades`.

```{r, include=FALSE}
# Create table with all the information we need
geo_trades <- trades %>%
  left_join(trades_capitals[,2:3], by = c("country_name" = "trades_countries")) %>%
  left_join(capitals, by = c("capitals_countries" = "country_name")) %>%
  filter(!is.na(capitals_countries)) %>%
  select(-capitals_countries, -country) %>%
  as_tibble()
```

## Creating the globe

To create the arcs in the globe we need a table in a very special format. It has to contain 4 columns: 
`origin_lat`, `origin_long`, `dest_lat`, and `dest_long` which will be the starting and ending points 
for each arc in the globe.

I'll use `geo_trades` to create a new table called `arcs` (our table will also have the value of the 
trades between the two points, but we'll exclude this column when passing the table to the globe-generating 
function). This is how I created `arcs`:

```{r}
# Create origins and destinations for arcs (origin is always Tokio)
arcs <- geo_trades %>%
  cbind(
    origin_lat = 35.68,
    origin_long = 139.69
  ) %>%
  select(-country_name) %>%
  transmute(
    origin_lat = origin_lat,
    origin_long = origin_long,
    dest_lat = lat,
    dest_long = long,
    value = value
  )
```

And now for the grand finale: this is the function we use to actually generate the globe. I encourage you 
to mess around with the parameters to see what each of them do, but feel free to just click and drag the 
globe around :D

```{r}
# Get the image of the globe
earth <- system.file("images/world.jpg",  package = "threejs")

# Create empty globe
globejs(
  img = earth, lat = arcs$dest_lat, long = arcs$dest_long, arcs = arcs[, 1:4],
  arcsOpacity = 0.6, arcsHeight = 0.8, arcsLwd = arcs$value,
  arcsColor = "green", atmosphere = TRUE, height = 800,
  width = 800, bg = "white", value = 4
)
```
