---
title: "Dissemination of Paid Vacations - Brazil's House of Deputies"
author: "Caio Lente"
date: "2017-02-12"
output: 
  html_document: 
    fig_height: 7
    fig_width: 9.3
---

```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(plotly)
library(tidyverse)
library(lubridate)
library(RColorBrewer)

# Read data
deputies <- read_csv("Data/dirty_deputies_v2.csv")

# Get summaryary of each type of refund
description_summary <- deputies %>%
  group_by(refund_description) %>%
  summarise(
    refund_cnt = n(),
    refund_avg = mean(refund_value),
    refund_tot = sum(refund_value)
  ) %>%
  ungroup()

# Get summaryary of each day
time_summary <- deputies %>%
  filter(!is.na(deputy_state)) %>%
  filter(!is.na(refund_date)) %>%
  mutate(refund_date = as_date(refund_date)) %>%
  group_by(refund_date, wday = wday(refund_date)) %>%
  summarise(
    refund_cnt = n(),
    refund_avg = mean(refund_value),
    refund_tot = sum(refund_value)
  ) %>%
  ungroup()
```

My goal with this analysis is to find out whether Brazilian deputies
have been using their reimbursement rights inadequately.

These reimbursements are called Quota for the Exercise of Parliamentary
Activity and are "designated to pay for expenses exclusively linked to
the exercise of parliamentary activity". Therefore, as far as I could tell,
there are two main ways we could detect an improper reimbursement claim
(given the available data):
* If the refund category is suspicious or
* If the time component of the refund is suspicious

## Dissemination of Parliamentary Activity

To investigate suspicious refund categories I tried plotting a bar chart
of total refunds per category (of the top 7 categories). Much to my surprise,
I had already found something very weird.

```{r reorder_descriptions, include=FALSE}
# Arrange descriptions according to total value
description_summary <- description_summary %>%
  arrange(refund_tot)

# Reorder deputies according to median
description_summary$refund_description <- ordered(
  description_summary$refund_description,
  levels = description_summary$refund_description
)
```

```{r bar_plot, warning=FALSE}
# Plot total value of refunds
description_summary %>%
  filter(
    row_number() > 10
  ) %>%
  plot_ly(
    x = ~refund_description,
    y = ~refund_tot,
    type = "bar",
    color = ~refund_description,
    colors = "Accent",
    opacity = 0.70
  ) %>%
  layout(
    legend = list(
      orientation = 'h',
      traceorder = "reversed",
      x = 0.5, y = -100
    ),
    yaxis = list(
      title = ""
    ),
    xaxis = list(
      title = "",
      showticklabels = FALSE
    )
  )
```

Apparently "dissemination of parliamentary activity" is the category
that has had the highest overall cost for the taxpayer: a total of
R$48,645,429.54. Since this refund description is very vague, it seems
to me that it is being widely used by the deputies as a cover up improper
refunds.

Just to make sure I wasn't being too quick to judge, I decided to look
into this a little further and created a box-plot for the 7 categories
with the highest mean refund values.

```{r reorder_deputies, include=FALSE}
# Get median of refunds for each category
d <- deputies %>%
  group_by(refund_description) %>%
  summarise(med = median(refund_value)) %>%
  arrange(med) %>%
  select(refund_description)

# Reorder deputies according to median
deputies$refund_description <- ordered(
  deputies$refund_description,
  levels = d$refund_description
)
```

```{r box_plots, warning=FALSE}
# Plot refund descriptions
deputies %>%
  filter(
    refund_description %in% d$refund_description[11:17]
  ) %>%
  plot_ly(
    x = ~refund_description,
    y = ~refund_value,
    type = "box",
    color = ~refund_description,
    colors = "Paired"
  ) %>%
  layout(
    legend = list(
      orientation = 'h',
      traceorder = "reversed",
      x = 0.5, y = -100
    ),
    yaxis = list(
      title = ""
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = "",
      mirror = TRUE
    )
  )
```

In the image above, dissemination of parliamentary activity doesn't have
the highest median, but it's outliers stand out from the rest. If we
examine to top outlier of this category (and of the whole plot), we find
that it corresponds to R$184,500.00 being reimbursed for expenses at a
*small print shop*, corroborating to the hypothesis that this category
is in fact being misused by the deputies.

The next logical step would be doing some text mining on the company names
to identify their types, and then check if they correspond with the assigned
refund categories. For the sake of brevity, I won't attempt to do this.

## Paid Vacation

Moving on to analysing the time component of the reimbursements, I decided
to get the average value of the refunds for each day of the year; with this
we can tell if there is any time frame where there is an unusual number of
reimbursement requests. For instance, finding out that there are too many
refunds of weekends would be suspicious.

After creating loads of visualizations (which you can reproduce by running
the code on my [Kaggle github repo](https://github.com/ctlente/Kaggle/tree/master/House)),
I thought there was something weird with the plot below. It represents the
density distribution of the mean value of refunds by day, see if you can
guess what grabbed my attention...

```{r get_densities, include=FALSE}
# Get density of mean refund (day by day)
t <- density(time_summary$refund_avg)
```

```{r plot_density, warning=FALSE, message=FALSE}
# Plot density of mean amount reimbursed (day by day)
plot_ly(x = t$x, y = t$y, mode = "lines", fill = "tozeroy")
```

See all those lumps after R$1.5K? It means that there is a large but
irregular number of days with average refund value higher than R$1.5K.
If there is a time pattern to these outliers, this could mean there
is something weird going on.

```{r plot_ts, warning=FALSE, message=FALSE}
# Plot mean value of refunds
time_summary %>%
  filter(refund_date > as_date("2016-01-01")) %>%
  plot_ly(
    x = ~refund_date,
    y = ~refund_avg,
    type = 'scatter',
    mode = 'lines',
    showlegend = FALSE
  ) %>%
  layout(
    yaxis = list(title = ""),
    xaxis = list(title = "")
  ) %>%
  add_trace(
    x = c(as_date("2016-12-23")),
    y = c(0, 10000),
    mode = "lines"
  )
```

And as expected, if we look at the time series of mean refund value
per day, we can clearly see that the days with the highest averages
are concentrated around late December and early January (especially
after 23/Dec, represented by the orange line, which is the day the
summer break starts).

In my opinion this time series shows that some deputies might be using
reimbursements to pay for their vacations.

## Conclusion

This wasn't in any way an exhaustive report, but I think it is already
able to supply enough evidence to the idea that Brazilian deputies are
using public money for personal gain. In the beginning we set out
to investigate two aspects of the dataset, and were able to find
suspicious activity on both; there are probably many other ways to
look at this data that I didn't think of or didn't have time for, so
I highly encourage you to analyse it for yourself!

If you'd like to know more about this subject I suggest
[Serenata de Amor](https://github.com/datasciencebr/serenata-de-amor),
an initiative created by Brazilian data scientists that uses AI to
flag suspicious reimbursement claims. And I also have a [blog](ctlente.com)
where I post regularly about my data visualization and analyses,
so if you liked this one there is a big chance you'll like the rest!
